Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-07-26 TEX/draft4.tex
Anything

\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{mathtools} 
\usepackage[margin=1.2in]{geometry}

@@

\newcommand{\reals}{\mathbb{R}}
\newcommand{\RL}{\mathbb{R}^L}
\newcommand{\CL}{\mathbb{C}^L}
\newcommand{\RN}{\mathbb{R}^N}
\newcommand{\RNN}{\mathbb{R}^{N\times N}}
\newcommand{\CNN}{\mathbb{C}^{N\times N}}
\newcommand{\inner}[1]{\left\langle {#1} \right\rangle}
\newcommand{\hx}{\hat{x}} 
\newcommand{\one}{\mathbf{1}} 
\newcommand{\SNR}{\ensuremath{\textsf{SNR}}}

@@

\TODO{Revise--Cryo--electron microscopy (cryo--EM) is an innovative technology for single particle reconstruction (SPR) of macromolecules.} 
%In recent years, structures of many molecules, previously regarded as insurmountable, are now being reconstructed to near-atomic resolution
%; see for instance~\cite{kuhlbrandt2014resolution,bartesaghi20152,smith2014beyond}.
%This technological advancement was recognized by the 2017 Nobel Prize in Chemistry~\cite{nobel}. 
% 
In a cryo--EM experiment, biological samples are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. The microscope produces a 2-D image of the samples embedded in the ice called a \emph{micrograph}. Each micrograph contains tomographic projections of the samples at unknown locations and under unknown viewing directions. The goal is to construct 3-D models of the molecules from the micrographs.

The signal to noise ratio ($\SNR$) of the projections in the micrographs is a function of two dominating factors. On the one hand, the $\SNR$ is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the $\SNR$ is a function of the molecule size. The smaller the molecules, the fewer detected electrons carry information about them.

%\paragraph{Particle picking.}
All contemporary methods in the field split the reconstruction procedure in several stages.
The first stage consists in extracting the various particle projections from the micrographs. This is called \emph{particle picking}. Later stages aim to construct a 3-D model of the molecule from these projections. The quality of the reconstruction eventually hinges on the quality of the particle picking stage. As can be seen from Figure~\ref{fig:micro_example}, locating the particles becomes increasingly challenging as the $\SNR$ degrades.

Crucially, it can be shown that reliable detection of individual particles is impossible below a certain critical $\SNR$. This fact has been recognized early on by the cryo-EM community. In particular, in a highly cited paper from 1995, Henderson~\cite{henderson1995limitations} investigates the following questions:
\begin{quote}
	\emph{For the purposes of this review, I would like to ask the question: what is the smallest size of free-standing molecule whose structure can in principle be determined by phase-contrast electron microscopy? Given what has already been demonstrated in published work, this reduces to the question: what is the smallest size of molecule for which it is possible to determine from images of unstained molecules the five parameters needed to define accurately its orientation (three parameters) and position (two parameters) so that averaging can be performed?}
\end{quote}
In that paper and in others that followed (e.g.,~\cite{glaeser1999electron}), it was established that particle picking is impossible for molecules below a certain weight.
%This has often been understood as meaning that particle reconstruction itself is impossible for such molecules~\cite{khoshouei2017cryo,liu2018nearatomic,liang2017phase}. \TODO{This list may be unfair, as some of these papers (in particular, \cite{liang2017phase}) just state that small molecules are very challenging, which is true.}
As a result, it is impossible to reconstruct such small molecules by any of the existing computational pipelines for single particle analysis, as the particles themselves cannot be picked from the micrographs.
%
%\TODO{either here, or in a close neighborhood of this paragraph you should mention the advances that the field has made in recent years towards reconstruction of small size molecules: mention Volta phase plate and the scaffolding cage (and cite the papers I sent you).}
This has motivated recent technical advances in the field, including the use of Volta phase plates~\cite{khoshouei2017cryo,liang2017phase} and scaffolding cages~\cite{liu2018nearatomic}.
%
%
Despite this progress, detecting small molecules (below 50 kDa) in the micrographs remains a challenge.
We note that nuclear magnetic resonance (NMR) spectroscopy and X-ray crystallography are well suited to reconstruct small molecules. Yet, cryo--EM has a lot to offer even for molecules with already known structures obtained via NMR spectroscopy or X-ray crystallography, because these methods have limited ability to distinguish conformational variability. \TODO{Need a ref for this claim.}

In this paper, we argue that there is a gap between the two questions in the quoted excerpt above, and that one may be able to exploit it to design better reconstruction algorithms.
Specifically, the impossibility of particle picking does not necessarily imply impossibility of particle reconstruction.
%This can be understood from a statistical point of view by counting the number of parameters that one aims to estimate. \TODO{For a given target resolution, the number of parameters required to describe the 3-D model of the molecule is fixed: it does not depend on the amount of data collected. In contrast, particle picking involves estimation of several so-called nuisance parameters for each projection present in the micrographs, such as their locations for example. To overcome low $\SNR$s, the ratio of data to parameters must increase. This is impossible if the number of parameters grows with the amount of data collected, but not so if the number of parameters remains fixed. Simply put, the key is to bypass the particle picking stage.}
%%Thus, acquiring more data does not necessarily allow to improve the particle picking stage, whereas it should in principle allow to improve the molecule reconstruction.
%%Indeed, the signal itself (the many different views of the molecule) is still present in the micrograph: 
%Indeed, in principle at least, the micrograph does contain the necessary information (several views of the molecule(s) of interest). While we may not be able to locate these views, we may still be able to extract the relevant information through other means.
Indeed, the aim is only to reconstruct the molecule: estimating the locations of the particles in the micrograph is merely a helpful intermediate stage when it can be done. Our main message is that the limits particle picking imposes on molecule size do not translate into limits on particle reconstruction.

%\TODO{Paragraph about the simplified model}
As a proof of concept, we study a simplified model where an unknown image appears numerous times at unknown locations in several micrographs, each affected by additive Gaussian noise---see Figure~\ref{fig:micro_example} for an illustration.
The goal is to estimate the planted image. The task is interesting in particular when the $\SNR$ is low enough that particle picking cannot be done reliably.
%Figure~\ref{fig:Einst_example} displays the results of our method, detailed in Section~\ref{sec:results}.
A precise mathematical formulation of the model, including an extension where more than one planted images are to be recovered, is provided in Section~\ref{sec:methods}.
To be clear, we do not consider here many prominent features of real SPR experiments and do not aim to reconstruct any 3-D structure. Instead, we solve a simpler problem that we believe captures key elements of the SPR problem. We note that similar models emerge in spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

In order to recover the planted image, we use autocorrelation analysis. In a nutshell, we relate the autocorrelation functions of the micrographs to the autocorrelation functions of the planted image.
%To do so, we assume the image occurrences are well separated.
For any noise level, these autocorrelations can be estimated to any desired accuracy, provided individual occurrences of the image are well separated and the image appears sufficiently many times in the micrographs. Importantly, there is no need to detect individual occurrences. The autocorrelations of the micrographs are straightforward to compute and require only one pass over the data. These directly yield estimates for the autocorrelations of the image. To estimate the image itself from its estimated autocorrelations, we solve a nonlinear inverse problem via least-squares or a phase retrieval algorithm; see Figure~\ref{fig:Einst_example} for an illustration and Section~\ref{sec:methods} for details. As a side note, we mention that expectation-maximization (EM)---a popular framework in SPR---is intractable for this problem; see Appendix~\ref{sec:theory} for a discussion.

Another interesting feature of the described approach pertains to model bias, whose importance in cryo-EM was stressed by a number of authors~\cite{vanheel1992correlation,shatsky2009method,henderson2013avoiding,vanheel2013finding}. In the classical ``Einstein from noise'' experiment, multiple realizations of pure noise are aligned to a picture of Einstein using cross-correlation and then averaged. In~\cite{shatsky2009method}, it was shown that the averaged noise rapidly becomes remarkably similar to the Einstein template. In the context of cryo--EM, this experiment exemplifies how prior assumptions about the particles may influence the reconstructed structure. This model bias is common to all particle picking methods based on template matching. In our approach, no templates are required, thus significantly reducing concerns about model bias.

%\TODO{Paragraph about bias; can also mention computational aspects (single pass, etc.)}

%\TODO{Already mention challenges? I think it's a good place.} -- no, not here: in conclusion

%\clearpage

@@

%Besides the model bias, particle pickers have several limitations due to the high noise level, in particular for small particles and in low contrast. 
%As a result, the projections in the 2-D images are typically not centered; including the correct centers as parameters dramatically increases the complexity of reconstruction algorithms.
%In addition, the information from particles that are too close to each other is usually neglected. Hence, valuable information that can be harnessed is omitted.

%\paragraph{Simplified model for SPR}

%In this paper, we present a preliminary study that aims to show that estimating a particle 
%directly from the micrograph, without identifying the location of individual particle images in the micrograph, might be possible.

%In particular, we consider the toy problem of estimating a set of signals $x_1,\ldots,x_K$ from their multiple occurrences in unknown  locations in a noisy data sequence $y$. Here, $y$ plays the role of the micrograph and the $K$ signals can be thought of as $K$ different viewing directions of the particle, each of which  appears multiple times in the micrograph. 
%A precise mathematical formulation of the model and the estimation problem is provided in Section~\ref{sec:methods}.
%To be clear, we do not consider here many prominent features of real SPR experiments and do not aim to reconstruct any 3-D structure; instead, we solve a simpler problem that we believe captures key elements of the true SPR problem. We also mention in passing that similar models emerge in different scientific fields, such as spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

\begin{figure}[t]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of micrographs of size $250\times 250$ with additive white Gaussian noise of variance $\sigma^2$ for increasing values of $\sigma$. Each micrograph contain the same four occurrences of a $50 \times 50$ image of Einstein. In panel (c), the noise level is such that it is very challenging to locate the occurrences of the planted image. In fact, it can be shown that at low $\SNR$, reliable detection of individual image occurrences is impossible, even if the true image is known. By analogy to cryo-EM, this depicts a scenario where particle picking cannot be done.}	
\end{figure}

%\paragraph{Autocorrelation analysis.}
%We now return to the general problem of estimating $x_1,\dots,x_K$ from the noisy observation $y$.
%In order to recover the signals $x_1,\dots,x_K$ from the data, we use autocorrelation analysis.
%In a nutshell, the method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by the maximum signal size and each signal appears sufficiently many times in the data. There is no need to detect individual occurrences.
%In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS) algorithm or a phase retrieval algorithm; see Section~\ref{sec:methods} for details.
%This method requires only one pass over the data and can recover the signals in any $\SNR$, if the signals appear enough times in the data. As a side note, we mention that expectation-maximization---a popular framework in SPR---is intractable for this problem (see Appendix~\ref{sec:theory}  for more details).

%\paragraph{Model bias.} The significance of the model bias in cryo--EM was stressed in~\cite{shatsky2009method,henderson2013avoiding} and demonstrated by the ``Einstein from noise'' example.
%In this experiment, the image of Einstein is correlated with multiple images of pure noise. The noisy images are aligned to Einstein's image using cross-correlation and then averaged. 
%In~\cite{shatsky2009method}, it was shown that with merely  1000 noise images, the averaged image is remarkably similar to the template image of Einstein rather than to pure noise.
%In Figure~\ref{fig:EinsteinFromNoise} we conducted this experiment with different number of noise images. With merely 1000 noise images, Einstein's face is clearly observed in the averaged image. With $10,000$  image, the output is remarkably similar to the template image of Einstein rather than to pure noise. Crucially, the more noisy images we use, the more severe the model bias towards the template. 
%In the context of cryo--EM, this example demonstrates how our prior assumptions about the particle may influence the reconstructed structure.
%This true in general to any method which is based on template matching.
%\TODO{Do we have a nice experiment to exemplify the phenomenon?}

\section{Results} \label{sec:results}

%\TODO{Results: What answer was found to the research question; what did the study find? Was the tested hypothesis true?}

%To support the claim that it may be possible to bypass particle picking in SPR from micrographs, as a proof of concept,
We conducted two experiments in the simplified image formation model described in the introduction:
\begin{enumerate}
	\item The first experiment aims to recover a 2-D image from an increasing number of micrographs with high noise, similar to the rightmost panel of Figure~\ref{fig:micro_example}. This is done using moments of second order, as these are sufficient to recover a 2-D image up to elementary symmetries;
	\item The second experiment aims to recover three distinct 1-D signals from an increasing number of 1-D micrographs with high noise. For this task, it is necessary to use moments up to third order.
\end{enumerate}
As outlined below, we find that it is indeed possible to recover good estimates of the ground truth signals from the highly corrupted micrographs, without particle picking. Furthermore, we find that the quality of estimation increases with the amount of data collected, despite the fact that particle picking remains challenging. The methods section provides additional details. In the discussion section, we outline how the general approach could be extended to full 3D SPR.

%In this section we describe  some numerical experiments and show results. The experimental details and the algorithms are discussed at length in Section~\ref{sec:methods}.

In the first experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. The latter are chosen so that two occurrences are always separated by at least 49 pixels. Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to $\SNR=1/20$). This high noise level is illustrated in the right panel of Figure~\ref{fig:micro_example}. In this first experiment, we assume knowledge of $\sigma$ and of the total number of signal occurrences across all micrographs.

We compute the average autocorrelation of the micrographs (equivalently, the average of their power spectra). This is a particularly simple computation. In the methods section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the power spectrum of the unknown image itself. Mathematically, it is easy to show that the quality of this estimate improves steadily as the amount of data grows, regardless of noise level. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{bauschke2004rrr,elser2017rrr}. RRR is initialized with a picture of the physicist Isaac Newton and iterates to produce the estimate, up to a reflection ambiguity.\footnote{\TODO{Initializing with a face is debatable: it might help the algorithm. On the other hand, I agree that it supports the notion that we don't get trapped too easily.}}

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs. (A movie is available in \TODO{supplementary material}.) Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available. Error is measured as the ratio of the root mean square error (RMSE) to the norm of the ground truth (square root of the sum of squared pixel intensities.) This is computed after fixing elementary symmetries (see methods.) As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.

%defined as 
%\begin{equation}
%\text{RMSE}  := \frac{\|x - \hat{x}\|_{\text{F}}}{\|x \|_{\text{F}}},
%\end{equation} 
%where $x$ and $\hat{x}$ are, respectively, the underlying and estimated image, and $\|\cdot\|_{\text{F}}$ stands for the Frobenius norm.  
%\TODO{We have a movie in the supplementary material.}

%%%% This is about bias, and it's imprecise; let's leave it out?
%As an initial guess, we picked an image of the physicist Issac Newton. If the algorithm was prone to model bias, we would expect to get as an output an image that resembles Newton, similarly to the ``Einstein from noise.'' However, the experiment exhibits the desired phenomenon: the more data we collect, the better the reconstruction quality.

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[scale=0.45]{reconstruction1_cropped}
		\caption{\small $P = 512$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[scale=0.45]{reconstruction10_cropped}
		\caption{\small $P = 512\times 10$}
	\end{subfigure}%

		\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[scale=0.45]{reconstruction100_cropped}
		\caption{ \small $P = 512\times 10^2$ }
	\end{subfigure}%
	\begin{subfigure}[h]{0.45\textwidth}
		\centering
		\includegraphics[scale=0.45]{reconstruction1000_cropped}
		\caption{\small $P = 512\times 10^3$}
	\end{subfigure}%

%	\begin{subfigure}[h]{0.33\textwidth}
%		\centering
%		\includegraphics[scale=0.33]{Einstein_progress_cropped}
%		\caption{\label{fig:err_curve}\small Error curve}
%	\end{subfigure}%

%\begin{figure}[h!]
%	\centering
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{Newton}
%		\caption{\label{fig:newton}Newton  (model)}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction2}
%		\caption{\small $P =1024$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction20}
%		\caption{$P = 1024\times 10$}
%	\end{subfigure}%
%	\begin{subfigure}[h]{0.25\textwidth}
%		\centering
%		\includegraphics[scale=0.3]{reconstruction200}
%		\caption{$P = 1024\times 100$}
%	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein from micrographs at noise level $\sigma = 3$ (see Figure~\ref{fig:micro_example}(c)). Averaged autocorrelations of the micrographs allow to estimate the power spectrum of the target image. This does not require particle picking. A phase retrieval algorithm (RRR) produces the estimates here shown, initialized with an image of the physicist Isaac Newton. Estimates are obtained from $P$ micrographs (growing across panels), each containing $700$ image occurrences on average.---\TODO{Will add image with $P=512\times 10^4$ and image of Newton.}
		% At a noise level of $\sigma = 3$, this amounts to an $\SNR$ of $1/20$.
	}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{Einstein_progress}
\caption{\label{fig:error_per_micro} Relative root mean square error of the estimate of Einstein's image as a function of the number of observed micrographs (logarithmic scale along both axes.)}
\end{figure}

@@

In the second experiment, three 1-D signals, each of length $L = 21$, appear at random locations in one long 1-D signal, which we call a micrograph by analogy. Any two occurrences are separated by at least 20 entries. The signals appear respectively about 30, 20 and 10 million times in a micrograph of length 12.3 billion. The micrograph is then contaminated with additive white Gaussian noise. This results in an $\SNR$ of about $1/9$, while about 10\% of the micrograph contains signal. Neither the number of occurrences nor the noise level $\sigma$ are known.

In the methods section, we detail how autocorrelations of the micrograph can be used to estimate weighted averages of the autocorrelations of the target signals. The individual signals and their relative densities are then estimated from autocorrelations up to order three by solving a nonlinear least-squares problem.

Figure~\ref{fig:1Dheterosignals} shows how the estimates improve as we see a larger and larger fraction of the micrograph (that is, as more and more data becomes available.) As is clear from the picture, despite the high noise level which would make it very challenging to locate the signal occurrences, the individual signals can be estimated accurately given enough data. Furthermore, the propensity of each signal can also be estimated.

%In order to estimate the signals, we computed the first three autocorrelation functions of the data and then estimated the signals and their corresponding $\gamma_i$ using a nonconvex least-squares. 
%As can be seen, given enough signal occurrences, we can estimate accurately the signals. 
%The estimation quality of the triangle signal is poorer than the other two signals, a phenomenon that is explained using Proposition~\ref{prop:uniqueness}.

>>>>

\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{verbatim}
\usepackage{float}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{multirow}
\usepackage{color}
\usepackage{url}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{caption}
\usepackage{mathtools} 
\usepackage[margin=1.2in]{geometry}

@@

\newcommand{\reals}{\mathbb{R}}
\newcommand{\RL}{\mathbb{R}^L}
\newcommand{\CL}{\mathbb{C}^L}
\newcommand{\RN}{\mathbb{R}^N}
\newcommand{\RNN}{\mathbb{R}^{N\times N}}
\newcommand{\CNN}{\mathbb{C}^{N\times N}}
\newcommand{\inner}[1]{\left\langle {#1} \right\rangle}
\newcommand{\hx}{\hat{x}} 
\newcommand{\one}{\mathbf{1}} 
\newcommand{\SNR}{{\textsf{SNR}}}

@@

Single particle reconstruction (SPR) using cryo--electron microscopy (cryo--EM) is an innovative technology for reconstructing the 3-D structure of macromolecules. In recent years, structures
of many molecules, previously regarded as insurmountable, are now being reconstructed to near-atomic resolution; see for instance~\cite{kuhlbrandt2014resolution,bartesaghi20152,smith2014beyond}. This technological advancement was recognized by the 2017 Nobel Prize in Chemistry~\cite{nobel}.

In a cryo--EM experiment, multiple biological samples of the (ideally) same molecule are rapidly frozen in a thin layer of vitreous ice. Within the ice, the molecules are randomly oriented and positioned. Then, the microscope produces a 2D tomographic projection image, called a \emph{micrograph}, of the multiple samples embedded in the ice. The micrograph is dominated by noise due to the small electron doses that
can be applied to the specimen without causing radiation damage.
The cryo--EM problem is to estimate the structure of the molecule from the micrograph (or, typically, several micrographs).

\paragraph{Particle picking.}
All contemporary methods in the field split the reconstruction procedure into several subroutines. 
The first step in the pipeline is  known as \emph{particle picking}, in which one identifies the locations of the 2-D tomographic projections of the samples embedded within the noisy micrograph. The output of an ideal particle picker is a series of 2-D images, each containing one centered tomographic projection associated with an unknown viewing direction and an in-plane rotation. This series of images 
is then used to estimate the structure.

Many automatic and semi-automatic methods for particle picking have been proposed, based on edge detection, template matching and deep learning; see for instance~\cite{harauz1989automatic,ogura2004automatic,zhu2016deep,frank1983automatic,scheres2015semi,heimowitz2018apple}. 
However, most of these procedures are prone to \emph{model bias}. For instance, in the popular framework of RELION~\cite{scheres2015semi}, the user manually marks hundreds of spots on the micrograph, believed to contain projections. 
Therefore, the algorithm's performance depends on the prior assumptions of the users about the particle's structure; the same holds true for deep learning based approaches which require constructing labeled sets of data.
Other methods use disks or differences of Gaussians as templates~\cite{langlois2014automated,voss2009dog}.
Nowadays, it is also still popular to pick particles manually. This method, while it exploits the researcher's experience, is both tedious and subject to model bias.

The significance of  model bias in cryo--EM was stressed in~\cite{shatsky2009method,henderson2013avoiding} and demonstrated by the ``Einstein from noise'' example.
In this experiment, the image of Einstein is correlated with multiple images of pure white noise. The noisy images are aligned to Einstein's image using cross-correlation and then averaged. 
In~\cite{shatsky2009method}, it was shown that with merely  1000 such noisy images, the averaged image is remarkably similar to the template image of Einstein rather than to pure noise.
%In Figure~\ref{fig:EinsteinFromNoise} we conducted this experiment with different number of noise images. With merely 1000 noise images, Einstein's face is clearly observed in the averaged image. With $10,000$  image, the output is remarkably similar to the template image of Einstein rather than to pure noise. Crucially, the more noisy images we use, the more severe the model bias towards the template. 
In the context of cryo--EM, this example demonstrates how our prior assumptions about the particle may influence the reconstructed structure.
This is true in general for any method which is based on template matching.
\TODO{TB: Do we have a nice experiment to exemplify the phenomenon?}

@@

Besides the model bias, particle pickers have several limitations due to the high noise level, in particular for small particles and in low contrast. 
As a result, the projections in the 2D images are typically not centered; including the correct centers as parameters dramatically increases the complexity of reconstruction algorithms.
%In addition, the information from particles that are too close to each other is usually neglected. Hence, valuable information that can be harnessed is omitted.

\paragraph{Simplified model for SPR.} In this paper, we present a preliminary study that aims to show that estimating a particle 
directly from the micrograph, without identifying the location of individual particle images in the micrograph, might be possible.

In particular, we consider the toy problem of estimating a set of signals $x_1,\ldots,x_K$ from their multiple occurrences in unknown  locations in a noisy data sequence $y$. Here, $y$ plays the role of the micrograph and the $K$ signals can be thought of as $K$ different viewing directions of the particle, each of which  appears multiple times in the micrograph. 
A precise mathematical formulation of the model and the estimation problem is provided in Section~\ref{sec:methods}.
To be clear, we do not consider here many prominent features of real SPR experiments and do not aim to reconstruct any 3-D structure; instead, we solve a simpler problem that we believe captures key elements of the true SPR problem. We also mention in passing that similar models emerge in different scientific fields, such as spike sorting~\cite{lewicki1998review}, passive radar~\cite{gogineni2017passive} and system identification~\cite{ljung1998system}.

Figure~\ref{fig:micro_example} shows a simple example of our problem with  one signal ($K=1$). 
Each panel demonstrates an image of size $250\times 250$, containing 4 occurrences of an Einstein image of size $50\times 50$. 
The left panel shows the data without noise. In the middle panel,  i.i.d.\ Gaussian noise with standard deviation $\sigma=0.5$  was added.  At this noise level, Einstein's locations are still easily identified. In this scenario it is likely that standard detection algorithms can locate the copies of the signal within the micrograph. In the right panel, we show the same data swamped in  
i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$.
Clearly, identifying the locations of Einstein in this micrograph is much more challenging. In fact, it can be shown that in the low signal--to--noise (\SNR) regime, 
detection  of individual image occurrences is impossible---even if the true image is known---and therefore particle picking is impossible.  This phenomenon is explained in detail in Appendix~\ref{sec:theory}. \TODO{WL: I think this example should go in the previous section on particle picking. The main issue we're trying to overcome is when detection is impossible; the fact that we don't use a template seems to be an additional advantage of our method, but not the main selling point. So I think the impossibility of detection in sufficiently low SNR must be emphasized earlier.}

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
RR	\caption{\label{fig:micro_example} Example of data sequences (micrographs) of size $250\times 250$ with additive i.i.d.\ Gaussian noise with variance $\sigma^2$. Each micrograph contains four occurrences of a $50 \times 50$ image of Einstein.}	
\end{figure}

\paragraph{Autocorrelation analysis.}
We now return to the general problem of estimating $x_1,\dots,x_K$ from the noisy observation $y$.
In order to recover the signals $x_1,\dots,x_K$ from the data, we use autocorrelation analysis.
The method consists of two stages. First, we estimate a mixture (i.e., linear combination) of the low--order autocorrelation functions of the signals from the data. These quantities can be estimated, to any desired accuracy, if individual occurrences are separated by the maximum signal size and each signal appears sufficiently many times in the data. There is no need to detect individual occurrences.
In the second stage, the signals are estimated from the mixed autocorrelations using a nonconvex least-squares (LS) algorithm or a phase retrieval algorithm; see Section~\ref{sec:methods} for details.
This method requires only one pass over the data and can recover the signals in any $\SNR$, if the signals appear enough times in the data. As a side note, we mention that expectation-maximization---a popular framework in SPR---is intractable for this problem (see Appendices~\ref{sec:related_literature} and~\ref{sec:theory}  for more details).

\section{Results}

In this section we describe  some numerical experiments and show results. The experimental details and the algorithms are discussed at length in Section~\ref{sec:methods}.

In the first experiment, we estimated Einstein's image of size $50\times 50$ from multiple micrographs of size $4096\times 4096$ pixels, contaminated  with additive i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ ($\SNR=1/20$). An illustrative example of a micrograph appears in the right panel of Figure~\ref{fig:micro_example}.
Each micrograph contains multiple occurrences of Einstein's image  in random locations, while keeping a separation of at least $49$ pixels between the images in each direction. From the micrographs, we estimated the signal's power spectrum, which is equivalent to the second-order autocorrelation. Then, in order to estimate the signal itself, we applied a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017matrix}.

As an initial guess, we picked an image of the physicist Issac Newton; see Figure~\ref{fig:newton}. If the algorithm was prone to model bias, we would expect to get as an output an image that resembles Newton, similar to the ``Einstein from noise'' effect. However, the experiment exhibits the desired phenomenon: the more data we collect, the better the reconstruction quality.

Figure~\ref{fig:Einst_example} demonstrates several recovery results for different number of recorded micrographs. Figure~\ref{fig:error_per_micro} presents the normalized recovery error and the power spectrum estimation error as a function of the number of micrographs.
To measure the error, we use the root mean square error (RMSE) defined as 
\begin{equation}
\text{RMSE}  := \frac{\|x - \hat{x}\|_{\text{F}}}{\|x \|_{\text{F}}},
\end{equation} 
where $x$ and $\hat{x}$ are, respectively, the underlying and estimated image, and $\|\cdot\|_{\text{F}}$ stands for the Frobenius norm. 
As expected, the RMSE of estimating the power spectrum   decreases linearly with slope $-1/2$ in logarithmic scale. 
\TODO{TB: Few comments: 1. We have a movie in the supplementary material. 2. Does the error also have the right slope? We'll know once we'll gather enough micrographs 3. Need to improve the micrograph's generation code  }

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{Newton}
		\caption{\label{fig:newton}Newton  (model)}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{reconstruction1}
		\caption{\small $P = 512$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{reconstruction10}
		\caption{$P = 512\times 10$}
	\end{subfigure}%

	\begin{subfigure}[h]{0.3\textwidth}
		\centering
		\includegraphics[scale=0.35]{reconstruction100}
		\caption{$P = 512\times 100$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.35]{reconstruction200}
	\caption{$P = 512\times 200$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.3\textwidth}
	\centering
	\includegraphics[scale=0.35]{reconstruction840}
	\caption{$P = 512\times 840$}
	\end{subfigure}%

	\caption{\label{fig:Einst_example} Recovery of Einstein's image. The RRR algorithm is initialized with the image of the physicist Issac Newton in the upper-left panel. We show estimations of Einstein using $P$ number of micrographs, each containing $700$ image occurrences on average with an $\SNR$ of $1/20$. The signal's estimation RMSE is  $0.5874, 0.3731, 0.2517, 0.2137, 0.1534$. The power spectrum estimation RMSE is $0.0381, 0.012, 0.0038, 0.0027, 0.0013$. }
\end{figure}

@@

Results from a similar experiment with three ($K=3$) one-dimensional signals are presented in Figure~\ref{fig:1Dheterosignals}.
 We define the ratio of the space occupied by the $i$th signal as 
\begin{equation}
\gamma_i = \frac{M_i L}{N},
\end{equation}
where $M_i$ is the number of times signal $x_i$  occurs, $L$ is the length of each signal and $N$ is the micrograph length.
These ratios $\gamma_i$ represent the {density} of the the signals in the data.
In the 1-D experiment, we do not assume knowledge of either the ratios $\gamma_i$ or the noise level $\sigma$.

In order to estimate the signals, we compute the first three autocorrelation functions of the data and then estimate the signals and their corresponding $\gamma_i$ using a nonconvex least-squares. 
As can be seen, given enough signal occurrences, we can estimate the signals accurately. 
The estimation quality of the triangle signal is poorer than the other two signals, a phenomenon that is explained using Proposition~\ref{prop:uniqueness}.
\TODO{TB: 1. Figure~\ref{fig:1Dheterosignals} will be replaced  with a ``progress figure'' (like we did for Einstien) for each signals and a plot of the recovery error for all three signals. 2. We may want to replace the triangle signal}
%We do not assume the knowledge of the noise level. 
%Crucially, starting from the point of $n\approx10^8$, the RMSE of the signal estimation decreases linearly with slope $-1/2$---the expected estimation rate of the autocorrelation functions---implying the stability of the recovery algorithm.

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{Einstein_progress}
\caption{\label{fig:error_per_micro} The left panel shows the RMSE of estimating the power spectrum of the signal from the data as a function of the number of  collected micrographs. As expected, the curve is linear with slope $-1/2$ in logarithmic scale. The right panel shows the RMSE  (in logarithmic scale) of recovering Einstein's image. \TODO{TB: The program continues running so these figures will be extended to the right (more micrographs).}}	
\end{figure}
